---
title: "Growth rate workshop"
author: "Andrew Irwin"
date: "2021-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(broom)
# library(performance)
# library(see)
my_theme = theme_bw() + theme(text = element_text(size = 18))
```

Notes for a Finkel-Irwin workshop on growth rate in April 2021.

## Goals

* Overview
* Measuring biomass
* Two point growth rate estimates
* Growth rate regressions with R 
  * One batch culture
  * One series of semi-continuous batch cultures
  * Multiple replicates of semi-continuous batch cultures
* Changes in growth rate
  * Lag phase
  * Stationary phase
* When and how much to dilute?
* Data organization recommendations
* How to use this method with your own data
* Exercises
* Creating simulated data

## Overview

Growth rates are fundamental to lab work on phytoplankton. When we acclimate a 
phytoplankton culture to a set of growth conditions, one of the ways we assess if
the culture is acclimated is our ability to measure a constant growth rate with 
high precision. If you estimate growth rate from just two observations of biomass separated 
by a day, you should expect fluctuations in your growth rate regardless of how careful
you are with your cultures, because they way you estimates the biomass in a culture
will include errors and these errors will propagate into your growth rate calculation.
Multiple observations of biomass are required to get a good estimate of growth rate. 
You still need to be careful, because many observations can be used to get a very precise
estimate of growth rate, even if the growth rate is changing. So you also need to 
plot your data and in particular the residuals to see if there is a pattern over time
in the difference between the observed biomass and your predictions of biomass.

## Measuring biomass

You can measure changes in phytoplankton biomass by counting cells, measuring chlorophyll
fluorescence, or measuring chlorophyll in your culture. (Of course you could measure 
carbon or other element content directly, but this is a much more time-consuming mesurement.)

All of these measurements have advantages and disadvantages and can be unsuitable 
if the cultures are not well acclimated.

Regardless of the measurement method you should always record the date and time of the
observation or when the sample was taken from the culture. Don't just include the "day" 
since the start of the experiment.

### Cell counting

Cell counts are a direct measure of cell abundance. Counting under the microscope is 
time-consuming and usually counts the smallest number of cells, possibly leading to 
uncertainty from small numbers of cells counted. Counting with the Coulter (or Z) counter 
usually gives counts of thousands or tens of thousands of cells, but if cells are very small, 
background noise can contaminate your cell counts. Counting with the flow cytometer 
has the advantage of counting large numbers and can distinguish particles with chlorophyll
from debris without chlorophyll.

If you count with a microscope you should count about 100 cells and record the volume
(regions in the haemocytometer or Sedgewick-Rafter chamber) in triplicate from each 
sample. The role of the triplicate is to show you that you can get approximately the same 
number of cells per volume in a sample, which gives you a "real-time" confirmation that you
have counted enough cells to get a good estimate of abundance. If cell abundances are very low
you may not be able to find this many cells and your technical replicates will let you know
if you have a good estimate of cell abundance.

### Chlorophyll fluorescence

Chlorophyll fluorescence is the fastest way to measure the biomass and track how it
changes from day to day. The quantity you get can be converted into cell counts or 
chlorophyll content if you create a calibration curve from simultaneous measurements of both
quantities over a sufficient range of concentrations.

Key considerations for chlorophyll fluorescence are that the fluorescence will change depending on the light treatment the cultures are grown in and the light environment immediately before
the fluorescence measurement is made. Chlorophyll per cell, and thus fluorescence, will
change with the time of day (relative to the light:dark period) and other growth conditions including irradiance, nutrient concentration, and even temperature. For this reason you should dark-adapt
your cells for 10-15 minutes before measuring their fluorescence and measure their fluorescence
at approximately the same time every day (e.g., a fixed number of hours after the incubator lights come on.)

### Chlorophyll content

Chlorophyll content is measured by extracting the chlorophyll from cells in a solvent
and quantifying the amount using fluorescence or absorbance. This solves the problem of 
needing to dark adapt the cells but the chlorophyll content of the cells still varies
with the time of day and so samples should be taken at the same time. Samples can be preserved and 
measured later in batches, but this is not helpful if your goal is to measure growth rate
and check the acclimation of your cells.


## Two point growth rate estimates

Acclimated phytoplankton that are not limited for resources will grow exponentially
at a rate $\mu$. This means that their biomass over time can be predicted by 
the equation
$$B(t) = B(0)e^{\mu t}.$$

In principle, two observations of biomass taken a known amount of time apart
is sufficient to estimate the growth rate. In practice, this is very unreliable
and does not permit you to estimate the uncertainty in your estimate of 
growth rate. The "two point" estimate of growth rate is 
$$\mu  = \frac{\log B(t_2) - \log B(t_1)}{t_2-t_1} = \frac{\log B(t)/B(0)}{t}.$$


## Creating simulated data

In these notes I will simulate data for analysis.
Columns in the dataset are

* experiment + treatment code (e.g., species, growth conditions), possibly split over several columns
* date and time of observation
* replicate (culture line)
* batch number
* biomass (any units, but consistent; not log transformed)
* dilution factor (dimensionless ratio: old culture volume / total new culture volume)

If one batch culture is started from the previous with a known dilution, create a separate line to record the dilution factor and record the time distinct from any cell count observations, so that it is obvious which counts are from before and after the dilution.

If the dilution is known, an extrapolated biomass will be computed and the batch numbers will be discarded. If the dilution factor is not known, a separate intercept will be estimated for each batch.

### Parameters

```{r}
set.seed(1) # to make the results reproducible
n_replicates <- 3
biomass_0 <- 1000 # mean biomass at starting time of batch culture
biomass_prop_s <- 0.075 # standard deviation of counts as proportion of true value
record_dilution <- TRUE 
growth_rate <- 0.8
start_date <- ymd_hms("2021-03-13 09:30:00")
days_between_counts <- 2
dilute_at <- 30000 # target biomass to dilute cultures
n_counts <- 20 # number of samples to collect in the experiment
```

### Simulate the data

```{r}
counts <- tibble( experiment = "A", 
                  date = rep(start_date, n_replicates),
                  batch = 1,
                  replicate = 1:n_replicates,
                  true_biomass = biomass_0 * rnorm(n_replicates, 1, biomass_prop_s),
                  dilution = NA,
                  observed_biomass = NA
                  )
```

Add data.

```{r}
for (i in 1:n_counts) {
  last_obs <- counts %>% slice_tail(n = n_replicates)
  new_obs <- last_obs %>% mutate( date = date + days_between_counts*86400 - 15*60*(!is.na(dilution)),
                                dilution = NA,
                                true_biomass = true_biomass * exp(growth_rate*days_between_counts),
                                observed_biomass = true_biomass * rnorm(n_replicates, 1, biomass_prop_s)
  )
  counts <- bind_rows(counts, new_obs)
  dilute <- max(new_obs$observed_biomass) > dilute_at
  if (dilute) {
    dilute_data <- new_obs %>% 
      mutate(date = date + 15*60,
             batch = batch + 1,
             dilution = biomass_0/observed_biomass,
             true_biomass = true_biomass * dilution,
             observed_biomass = NA)
    counts <- bind_rows(counts, dilute_data)
  }
}
```

Add time in days and extrapolated biomass.

```{r}
counts <- counts %>% mutate(time_d = as.numeric(date - min(date))/86400) %>%
  replace_na(list(dilution=1)) %>%
  group_by(replicate) %>%
  mutate(extrapolated_biomass =  observed_biomass / cumprod(dilution) ) %>%
  ungroup()
```

## Growth rate regressions with R 

A much better way to estimate growth rate is from a series of 4 or more observations
of biomass and linear regression.  The equation is the same, but now you can estimate
the uncertainty in your estimate. We will organize our data in a standard way and 
use R to compute growth rate and diagnose if the biomass estimates indicate that
cells are or are not growing at a constant rate.

### One batch culture

Here is a graph of possible cell counts from a single batch culture.

```{r}
counts %>% filter(batch == 1, replicate == 1) -> ss1
ss1 %>%
  ggplot(aes(date, observed_biomass)) + geom_point() +
  scale_y_log10() + my_theme
```

Compute two-point growth rates for each pair of observations one day apart.

```{r}
ss1 %>% filter(!is.na(observed_biomass)) %>%
  mutate(growth_rate = log(observed_biomass / lag(observed_biomass))/(time_d - lag(time_d))) %>%
  select(date, time_d, growth_rate)
```

Compute a linear regression through these three points.

```{r}
ss1 %>% lm(log(observed_biomass) ~ time_d, data = .) %>% 
  tidy(conf.int = TRUE)
```

### One series of semi-continuous batch cultures

```{r}
counts %>% filter(replicate == 1) -> ss2
ss2 %>%
  ggplot(aes(date, extrapolated_biomass)) + geom_point() +
  scale_y_log10() + my_theme
```

Compute two-point growth rates for each pair of observations one day apart.

```{r}
ss2 %>% filter(!is.na(extrapolated_biomass)) %>%
  mutate(growth_rate = log(extrapolated_biomass / lag(extrapolated_biomass))/(time_d - lag(time_d))) %>%
  select(date, time_d, growth_rate)
```

Compute a linear regression through these three points.

```{r}
ss2 %>% lm(log(extrapolated_biomass) ~ time_d, data = .) -> m2
m2 %>% 
  tidy(conf.int = TRUE)
```

### Multiple replicates of semi-continuous batch cultures

One batch, with replicates.

```{r}
counts %>% filter(batch == 1) -> ss3
ss3 %>%
  ggplot(aes(date, observed_biomass,
             color = factor(replicate))) + geom_point() +
  scale_y_log10() + my_theme
```

Compute two-point growth rates for each pair of observations one day apart, averaging biomass
from the three replicates.

```{r}
ss3 %>% filter(!is.na(observed_biomass)) %>%
  group_by(date, time_d) %>%
  summarize(observed_biomass = mean(observed_biomass)) %>%
  ungroup() %>%
  mutate(growth_rate = log(observed_biomass / lag(observed_biomass))/(time_d - lag(time_d))) %>%
  select(date, time_d, growth_rate)
```
Compute a linear regression, with a single slope, but a different intercept for each 
replicate.

```{r}
ss3 %>% lm(log(observed_biomass) ~ time_d + factor(replicate), data = .) %>% 
  tidy(conf.int = TRUE)
```

All batches, with replicates.

```{r}
counts %>% lm(log(extrapolated_biomass) ~ time_d + factor(replicate), data = .) %>% 
  tidy(conf.int = TRUE)
```

## Changes in growth rate

### Lag phase

If you initiate a culture from a cells in stationary phase, for example from a 
culture that has been in "maintenance mode" and not regularly provided fresh nutrients
or reduced to low biomass, you should expect to see a "lag phase" in your growth data.
This can also happen with a culture that is growing well if conditions, such as nutrient
concentrations in the media are changed.

The most likely explanation for the lag phase is that cells are not well acclimated 
to the new conditions and so they do not grow well. With an "old" culture it is possible that
many of the cells you count are not viable at all. We use this idea to simulate a lag phase.
We start a culture with a specified cell concentration, but make a proportion of them 
unable to grow. The dead cells will gradually get diluted out as you initiate new cultures
by serial dilutions.

Calculations...




### Stationary phase

We have only simulated exponential growth. In a real experiment, cell growth will slow
down once cell density exceeds a threshold. There are many reasons: high pH / low DIC, 
light attenuation in the culture, and nutrient depletion. We don't model any 
resources here, so we will simulate this slow down in growth by establishing a carrying capacity.
The important feature you will see is that growth rate starts to slow down long before the 
cell density reaches the carrying capacity.


The solution to the logistic growth equation, with a carrying capacity $K$, is 
$$P(t) = P(0)\frac{Ke^{\mu t}}{(K-P(0))+P(0)e^{\mu t}}$$




* When and how much to dilute?
* Data organization recommendations
* How to use this method with your own data
* Exercises
* Creating simulated data

