---
title: "Liefer growth calculation"
author: "Andrew Irwin"
date: "19/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(janitor)
liefer <- read_csv("liefer-growth-data.csv") %>% clean_names()
```

Compute cell numbers, accounting for dilution. Plot cell numbers over time.

```{r}
liefer2 <- liefer %>% replace_na(list(dilution_factor = 1)) %>%
  group_by(species, replicate) %>%
  mutate(total_dilution = cumprod(dilution_factor),
         extrapolated_cell_density = cell_density / total_dilution)
liefer2 %>% ggplot(aes(days_in_n_free_media, extrapolated_cell_density)) +
  geom_point() + 
  scale_y_log10() + 
  facet_wrap(~ species, scales = "free_y")
```

Compute and plot two-point growth rates.

```{r}
my_mean <- function(x,y) {
 case_when( is.na(x) ~ y,
            is.na(y) ~ x,
            TRUE ~ (x+y)/2)
}
liefer3 <- liefer2 %>% 
  filter(!is.na(extrapolated_cell_density)) %>%
  mutate(growth_rate_fw = log(extrapolated_cell_density/lag(extrapolated_cell_density))/
           (days_in_n_free_media - lag(days_in_n_free_media)),
         growth_rate_bk = log(lead(extrapolated_cell_density)/extrapolated_cell_density)/
           (lead(days_in_n_free_media) - days_in_n_free_media),
         growth_rate_cn = my_mean(growth_rate_fw, growth_rate_bk))
liefer3 %>% ggplot(aes(days_in_n_free_media, growth_rate_cn)) +
  geom_point() + 
  facet_wrap(~ species, scales = "free_y")
```

Estimate maximum growth rate from noisy growth rates and cell quotas. The Droop model for growth rate is 
$$\mu = \mu_{max}\left(1-\frac{Q_{min}}{Q}\right)$$

We can estimate $Q_{min}$ and $\mu_{max}$ by linear regression with $\mu$ as the response 
and $1/Q$ as the predictor. The intercept will be the maximum growth rate and the slope is $-\mu_{max}Q_{min}$.

```{r}
liefer3 %>% ggplot(aes(1/n, growth_rate_cn)) +
  geom_point() + 
  facet_wrap(~ species, scales = "free")
```

This model can be fit for all species simultaneously using `lm` as shown in the next code block, but it takes a bit of effort to interpret the coefficients. 

```{r eval=FALSE}
m1 <- lm(growth_rate_cn ~ species*replicate + species:nRecip, 
         data = liefer3 %>% mutate(nRecip = 1/n))
tidy(m1)
coefplot::coefplot(m1)
```

Instead I will fit four models, one for each species. 

```{r}
library(broom.mixed)
library(lme4)
m2 <- liefer3 %>% 
  group_by(species) %>%
  nest() %>%
  mutate(model = map(data, ~ lmer(growth_rate_cn ~ (1|replicate) + I(1/n), data = .)),
         tidied = map(model, tidy, conf.int=TRUE)) %>%
  unnest(tidied) %>%
  select(-data, -model) 
```

Discard the differences in interept for replicates B and C 

Use Droop differential equation to estimate growth rate.
